Tulai

### 2023-09-18 paper/scraps/scraps-tulai.qmd
edited and polished the codes. Good to render now.

>>>>

PCA_al:

```{r}
#PCA: area_layer (al)
#I think as we are going to calculate PCA to understand other factors that might be affected the duration of the occupation, it would be logical to calculate it for each layer, instead of level. we have 5 stratigraphical layers: e, d, c, b, a
#d1:1.5*1.5/ ta: 4*5/ tp1:2*1.8
#e: excavated area is A/ depth of excavation is 0-75(mean)
#d: excavated area is A/ 0.75m-3m→ 2.25
#c: d1 and TP → 30 cm and 40cm
#b: TP/ 170 cm 
#a: D1: 70cm

# Add 'dimension' columns and relocate it after 'al' (based on Hole (1974))
library(dplyr)

summary_table_al_pca <- summary_table_al %>%
  mutate(
    length = case_when(
      al == "D1_a" ~ 1.5,
      al == "D1_c" ~ 1.5,
      al == "TA_d" ~ 4,
      al == "TP1_c" ~ 2,
      al == "TA_e" ~ 4,
      TRUE ~ NA_real_
    ),
    width = case_when(
      al == "D1_a" ~ 1.5,
      al == "D1_c" ~ 1.5,
      al == "TA_d" ~ 5,
      al == "TP1_c" ~ 1.8,
       al == "TA_e" ~ 5,
      TRUE ~ NA_real_
    ),
    thickness = case_when(
      al == "D1_a" ~ 0.7,
      al == "D1_c" ~ 0.3,
      al == "TA_d" ~ 2.25,
      al == "TP1_c" ~ 0.4,
      al == "TA_e" ~ 0.75,
      TRUE ~ NA_real_)
  ) %>%

  mutate(volume = length * width * thickness) %>%
  relocate(length, width, thickness, volume, .after = al)
# Show the updated data frame
summary_table_al_pca

```{r}
# Load required library
library(dplyr)

# Add a new column called 'blank_count'
summary_table_al_pca <- summary_table_al_pca %>%
  mutate(blank_count = blade_count + flake_count + bladelet_count) %>%
  select(al, length, width, thickness, volume, blade_count, flake_count, bladelet_count, blank_count, everything())

# Calculate core frequency, blank frequency, retouch frequency, lithic density 
summary_table_al_pca <- summary_table_al_pca %>%
  mutate(
    lithic_density = round((core_count + blank_count + core_rejuvenation_count) / volume, 2),
    core_freq = round(core_count / volume, 2),
    blank_freq = round(blank_count / volume, 2),
    retouch_freq = round(retouched_tool_count / volume, 2)
  )

# Calculate tool diversity using Menhinick's index
summary_table_al_with_diversity <- tl_final %>% 
  filter(retouch == 1) %>%
  group_by(al, `tool-typology`) %>%  
  summarise(n = n(), .groups = 'drop') %>% 
  group_by(al) %>%  
  summarise(
    num_unique_tool_types = n_distinct(`tool-typology`),
    total_retouched_tools = sum(n),
    tool_diversity = round(num_unique_tool_types / sqrt(total_retouched_tools), 2),
    .groups = 'drop'
  )

# Merge this new information into the existing summary_table_al_pca
summary_table_al_pca <- left_join(summary_table_al_pca, summary_table_al_with_diversity, by = "al")

# Remove the 'Total' row and select only the columns needed for PCA
fin_summary_table_al_pca <- summary_table_al_pca %>%
  filter(al != "Total") %>%
  select(al, core_freq, blank_freq, lithic_density, retouch_freq, tool_diversity)

# Show the final table
print(fin_summary_table_al_pca)

```{r}
library(FactoMineR)
# Perform PCA
res.pca <- PCA(fin_summary_table_al_pca[, -1], graph = FALSE)  # Exclude the 'al' column

```{r}
## Got error: couldnt figure out the reason of error

# Create a data frame to hold the rounded eigenvalues and the rounded percentage variance
eigen_df <- data.frame(
  Dimension = paste("Dim", 1:nrow(eigenvalues), sep = "."),
  Eigenvalue = round(eigenvalues[, 1], 3),
  `Variance percent` = round(eigenvalues[, 2], 3),
  `Cumulative variance percent` = round(cumsum(eigenvalues[, 2]), 3)
)

# Print the table
print(eigen_df)

```{r}
# variables to PC1
fviz_contrib(res.pca,  
             choice = "var",  
             axes = 1,  
             top = 10)
```

```{r}
fviz_contrib(res.pca,  
             choice = "var",  
             axes = 2,  
             top = 10)

```{r}
fviz_contrib(res.pca,  
             choice = "var",  
             axes = 3,  
             top = 10)
```

```{r}
fviz_contrib(res.pca,  
             choice = "var",  
             axes = 4,  
             top = 10)
```

```{r}
#Plot: Contribution of variables for each of the four relevant PCA dimensions
library(factoextra)
library(ggplot2)

#save contribution plots into one objects
p1 <- fviz_contrib(res.pca, 
                   choice = "var", 
                   axes = 1, 
                   top = 10)
p2 <- fviz_contrib(res.pca, 
                   choice = "var",
                   axes = 2, 
                   top = 10)
p3 <- fviz_contrib(res.pca, 
                   choice = "var", 
                   axes = 3, 
                   top = 10)
p4 <- fviz_contrib(res.pca, 
                   choice = "var", 
                   axes = 4, 
                   top = 10)

#combine individual plots into one plot 
library(gridExtra)
grid.arrange(p1, p2, p3, p4, ncol = 2)

```{r}
# Extract contributions of variables to each principal component
contributions <- res.pca$var$contrib

# Compute average contribution for each dimension
avg_contrib_dim1 <- mean(contributions[,1], na.rm = TRUE)
avg_contrib_dim2 <- mean(contributions[,2], na.rm = TRUE)
avg_contrib_dim3 <- mean(contributions[,3], na.rm = TRUE)
avg_contrib_dim4 <- mean(contributions[,4], na.rm = TRUE)

# Find variables that contribute more than average to each dimension
high_contrib_dim1 <- contributions[contributions[,1] > avg_contrib_dim1, 1]
high_contrib_dim2 <- contributions[contributions[,2] > avg_contrib_dim2, 2]
high_contrib_dim3 <- contributions[contributions[,3] > avg_contrib_dim3, 3]
high_contrib_dim4 <- contributions[contributions[,4] > avg_contrib_dim4, 4]

# Sort them in descending order
sorted_high_contrib_dim1 <- sort(high_contrib_dim1, decreasing = TRUE)
sorted_high_contrib_dim2 <- sort(high_contrib_dim2, decreasing = TRUE)
sorted_high_contrib_dim3 <- sort(high_contrib_dim3, decreasing = TRUE)
sorted_high_contrib_dim4 <- sort(high_contrib_dim4, decreasing = TRUE)

# Print or use for further analysis
print("Significant contributions to Dim1:")
print(sorted_high_contrib_dim1)

print("Significant contributions to Dim2:")
print(sorted_high_contrib_dim2)

print("Significant contributions to Dim3:")
print(sorted_high_contrib_dim3)

print("Significant contributions to Dim4:")
print(sorted_high_contrib_dim4)

```{r}
# screeplot; inspect distribution of PCs
library(factoextra)  
fviz_screeplot(res.pca)
```

```{r}
# Inspect eigenvalues
print(res.pca$eig)

```{r}
# Load the required package
library(kableExtra)

# Create a table with kable and kableExtra
eigen_df %>%
  kable("html") %>%
  kable_styling("striped", full_width = F) %>%
  column_spec(1, bold = T)
```

# Rounding the numerical columns to 3 decimal places
eigen_df[, -1] <- round(eigen_df[, -1], 3)

# Replacing 'comp' with 'Dim'
eigen_df$Dimension <- gsub("comp", "Dim", eigen_df$Dimension)

# Removing row names
rownames(eigen_df) <- NULL

# Creating the kable output
kable_output <- kable(eigen_df, 
                      "html", 
                      align = 'c', 
                      col.names = c("Dimension", 
                                    "Eigenvalue", 
                                    "Variance percent", 
                                    "Cumulative variance percent")) %>%
  kable_styling("striped", 
                full_width = F) %>%
  add_header_above(c(" " = 1, "Eigenvalues and percentage of variance for each dimension of PCA" = 3))

htmltools::browsable(kable_output)

```{r}
# Visualize biplot:1
fviz_pca_biplot(res.pca,
                axes = c(1, 2))
rownames(res.pca$ind$coord) <- fin_summary_table_al_pca$al
# Visualize biplot with row names
fviz_pca_biplot(res.pca)

```{r}
# Visualize biplot: 2
fviz_pca_biplot(res.pca,
                axes = c(3, 4))
rownames(res.pca$ind$coord) <- fin_summary_table_al_pca$al
# Visualize biplot with row names
fviz_pca_biplot(res.pca)

print(fviz_pca_biplot(res.pca, axes = c(3, 4)))

```{r}
# BM: My preference would be to show the calculation of these 
  # values from the raw data, rather than type them in 
  # here, where we can't see where they came from, we can't easily
  # verify these values as they are currently 
##SS: fixed

# Define the volumes for each area

# add the dimensions for each area
dimensions <- data.frame(
  area = c("ta", "tp1", "d1"),
  length = c(4, 2, 1.5),
  width = c(5, 1.8, 1.5),
  thickness = c(3, 1.6, 0.95)
)

# Calculate volume for each area based on the dimensions
dimensions <- dimensions %>%
  mutate(volume = length * width * thickness)

# Join artifact_counts with dimensions (which now includes volume) by area
lithic_density_df <- left_join(artifact_counts, dimensions, by = "area")

# Create the summary table
summary_table2_area <- tl_final %>%
  filter(area %in% c("ta", "tp1", "d1")) %>%
  group_by(area) %>%
  summarise(
    numb_retouched = sum(retouch == 1, na.rm = TRUE),
    total_lithics = n(),
    percent_retouch = round((numb_retouched / total_lithics), 2)
  )

kable(summary_table2_area, 
      caption = "Table 2. Lithic counts for each of the areas in Tula'i", 
      align = 'c',
      col.names = c("Area", 
                    "Number of Retouched Lithics", 
                    "Total Number of Lithics", 
                    "Retouch frequency", 
                    "Lithic Volumetric Density(pcs/(m^3)")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), font_size = 11) %>%
  column_spec(1, width = "1cm") %>% 
  column_spec(2, width = "1cm") %>%

------------------------------
Done with the main plots
------------------------------

```{r}
# WABI_LEVEL

# BM: I am confused why we are getting the counts of artefacts 
# and retouch pieces from the google sheet, after you wrote so many lines of
# code above to prepare the data from the CSV file. Where did the lithic data
# in the google sheet come from? My preference is to have 
# only copy of the lithic data in one place, and then join it with 
# other data using code. Currently it seems like we have the 
# lithic data in both the google sheet and the CSV file here. I'm worried that 
# will be confusing and be difficult to update and keep correct. I prefer that we 
# omit the lithic data from the google sheet, then import the google sheet and 
# join it to the lithic data from the CSV file. Then we can have a data frame
# with the area/layer/level data and lithic data to work on for plotting.

#SS: the googlesheet is the last thing I created. I created the summary table in R, and then for splitting up the depth (we have few depth range that was not possible to assign the layer to them, so, we splited them), I found it is easier to do so in Google sheet.

WABI_Level <- ggplot(filtered_data, 
            aes(x = lithic_dens,
                y = retouch_freq)) +
  geom_point(color = "black", size = 2) +
  geom_smooth(method = "lm", 
              se = TRUE, 
              color = "blue", 
              size = 0.5) +
  geom_text(aes(label = area_layer_level), 
            vjust = 1.5, 
            hjust = 0.5, 
            size = 2,
            check_overlap = TRUE) +
  stat_poly_eq(aes(label = paste(stat(eq.label), 
                                 stat(rr.label), 
                                 stat(p.value.label), 
                                 sep = "~~~")),
               formula = y ~ x, 
               parse = TRUE, 
               size = 2) +
  # Add titles and labels
  labs(
    title = "Relationship between Retouch Frequency and Lithic Volumetric Density",
    x = "Lithic Volumetric Density",
    y = "Retouch Frequency"
  ) +

WABI_Level

```{r}
# whole assemblage

  # Summarize data by Area
summary_table_tech <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  group_by(area) %>% 
  summarise(
   blade_count = sum(blank == "blade", na.rm = TRUE),
    flake_count = sum(blank == "flake", na.rm = TRUE),
    bladelet_count = sum(blank == "bladelet", na.rm = TRUE),
    bullet_core_count = sum(`core-typology` == "bullet", na.rm = TRUE),
    flat_core_count = sum(`core-typology` == "flat", na.rm = TRUE),
    pyramid_core_count = sum(`core-typology` == "pyramid", na.rm = TRUE),
    cylinder_core_count = sum(`core-typology` %in% c("prismatic", "cylinder"), na.rm = TRUE),
    shapeless_core_count = sum(`core-typology` == "shapeless", na.rm = TRUE),

```{r}
# plot_blank_type

# Data filtering and plotting
tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>% 
  filter(blank %in% c('bladelet', 'blade', 'flake')) %>%
  ggplot(aes(x = area, fill = blank)) +
  geom_bar(position = "dodge") +
  labs(
    x = "Area",
    y = "Number",
    fill = "Blank Type"
  ) +
  theme_minimal()

tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
  filter(!is.na(corefrag)) %>% 
  select(layer, corefrag) %>%
  ggplot() +
  aes(x = corefrag) +
  geom_bar() +  
  labs(x = "Core and Core fragments", y = "Frequency") +
  facet_wrap(~layer, ncol = 1) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

filtered_data <- tl_final %>%
  filter(!is.na(`tool-typology`) & `tool-typology` != "" & `tool-typology` != "NA")
ggplot(data = filtered_data) + 
  aes(y = `tool-typology`) + 
  geom_bar() +
  labs(x = "Frequency",
       y = "Tool Type") +
  theme_minimal() +
  theme(axis.text.y = element_text(angle = 20, hjust = 1, vjust = 0.5))

```{r}
# prepare the data for PCA-layer

# Calculating Blank Frequency
blank_freq <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(blank)) %>%
  group_by(layer) %>%
  summarise(blankFreq = sum(blank %in% c("bladelet", "blade", "flake")) / sum(!is.na(blank)), .groups = 'drop')

# Calculating Core Frequency
core_freq <- tl_final %>%
  filter(area %in% c("ta", "d1", "tp1")) %>%
  filter(layer %in% c("a", "c", "d", "e", "eor d"), !is.na(corefrag)) %>%
  group_by(layer) %>%
  summarise(coreFreq = sum(corefrag %in% c("core fragment", "bullet", "flat", "pyramid", "cylinder/prismatic", "core rejuvenation piece", "shapeless", "core on flake", "bipolar")) / sum(!is.na(corefrag)), .groups = 'drop')

# Printing the final table
final_table

#------------------------------------------------------------------------------
# BM: I wasn't able to run any code below this line
##SS: fixed
# -----------------------------------------------------------------------------

# Calculating Retouch Frequency
retouch_freq <- tl_final %>%
  filter(layer %in% c("a", "c", "d", "e", "eor d")) %>%
  group_by(layer) %>%
  summarise(retFreq = sum(retouch == 1, na.rm = TRUE) / n(), .groups = 'drop')

final_table

# Define estimated area and feature frequency data
area_data <- tibble(
  area = c("d1", "ta", "tp1"),
  feature_freq = c(1, 1, 0)
)

pca_final_table_area <- left_join(dimensions, retouch_freq, by = "area") %>%
  left_join(blank_freq, by = "area") %>%
  left_join(core_freq, by = "area") %>%
  left_join(lithic_density, by = "area") %>%
  left_join(tool_div, by = "area") %>%
  select(area, new_dimensions, feature_freq, retouchFreq, blankFreq, coreFreq, lithic_density, toolDiv)

# Define the values manually (you would get these from your PCA results)
# BM: I prefer not to write values in the code like this. I think
# this is bad practice and leads to mistakes
eigenvalues <- c(4.411234, 2.588766)
prop_var_explained <- c(63.02, 36.98)
cumulative_var_explained <- c(63.02, 100.00)
